---
title: "Simulations In-Class Project - DC"
date: "Due October 13, 2017 at 11:59pm"
output:
  html_document


---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```




#Project Goals:


With this project we will simulate a famous probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 

 1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
<<<<<<< HEAD
 
```{r}
set.seed(1)
Gamble = function(bet, bankroll, p){
  x = 0
  while(bet > 0) {
    simulation = runif(1)
    if(simulation <= p) {
      bankroll = bet + bankroll
      bet = (bet + bankroll) * 0.1
    } else {
      bankroll = bankroll - bet
      bet = bet
    }
    x = x+1
    if(bankroll < bet){
      break
      }
  }
  duration = x
  return(duration)
}

games = numeric()
s = 5000
for(i in 1:s)
{
  hand = Gamble(100,900,0.5)
  games = c(games, hand)
}
``` 
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet.
```{r}
a=length(games)
b=length(which(games <= 100))
prob = b/a
prob
``` 
    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
```{r}
a=length(games)
b=length(which(games <= 500))

prob=b/a
prob
``` 
    c. the mean time you go bust, given that you go bust within the first 5000 hands.
```{r}
a = games[which(games<=5000)]

mean=mean(a)
mean
``` 
    d. the mean and variance of your bankroll after 100 hands (including busts).
```{r}
set.seed(1)
Gamble1 = function(bet,bankroll,p){
  x = 0
  while(bet > 0) {
    simulation = runif(1)
    if(simulation <= p){
      bankroll = bet + bankroll
      bet = (bet + bankroll)*0.1
    } else {
      bankroll = bankroll - bet
      bet = bet
      }
    x = x+1
    if(x ==100){
      break
      }
    if(bankroll < bet){
      break
      }
  }
  duration = x
  return(bankroll)
  
}

bankroll = numeric()
s = 5000
for(i in 1:s)
{
  next.bet = Gamble1(100,900,0.5)
  bankroll = c(bankroll, next.bet)
=======
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
```{r}
## open libraries 
library(tidyverse)
## set seed for replicability, running the simulation 
set.seed(1234)

## function for gambler's ruin 
gamb.ruin <- function(bet, bankroll, p) {
  og.bet <- bet
  x <- 0 #empty vector to count number of games 
  while(bankroll >= bet) { ##run as long as you have money to bet
    game <- runif(1)
    if(game <= p) { ## if win, bet twice as much 
      bankroll = bankroll + bet
      bet = (bankroll+bet)*.10
    } else { ## if lose, bet back to original arguement 
      bankroll = bankroll - bet
      bet = og.bet
    }
  x <- x+1
    if(bankroll < bet){break} ##end if you don't have enough to bet 
  }
  times.played <- x 
  return(times.played)
}
```

```{r}
set.seed(1234)
num.games <- rep(NA, length = n) ##create empty numeric vector 
simulation <- 5000
for(i in 1:simulation) {
  hand <- gamb.ruin(bet=100, bankroll=900, p=.5) ##already betting 100, so you only have 900 to really play, place number of games in vector 
  num.games[i] <- hand
}

a <- length(num.games) #simualted games 
b <- length(
  which(num.games<=100) #amount of simulations that ended at/before 100 hands
)
a
b

PR <- b/a
PR
```
    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
```{r}
a <- length(num.games)
b <- length(
  which(num.games <=500)
)

PR <- b/a
PR
```
    c. the mean time you go bust, given that you go bust within the first 5000 hands.
```{r}
mean(
  num.games[which(num.games<=5000)]
)
```
    d. the mean and variance of your bankroll after 100 hands (including busts).
```{r}
## set seed for replicability, running the simulation 
set.seed(1234)
## function for gambler's ruin but w/ 100 games
gamb.ruin.bankroll <- function(bet, bankroll, p) {
  x <- 0
  og.bet <- bet 
  while(bankroll >= bet) { 
    game <- runif(1)
    if(game <= p) {  
      bankroll = bankroll + bet
      bet = (bankroll + bet)*.10
    } else { 
      bankroll = bankroll - bet
      bet = og.bet
    }
    if(bankroll < bet){break} 
    if(x == 100){break} ##end if hit 100 games 
  }
  x <- x+1
  times.played <- x 
  return(bankroll) ## return bankroll instead of num of games
}

bankroll <- rep(NA, length = n) ## set up empty bankroll 
simulation <- 5000
for(i in 1:simulation){
  hand <- gamb.ruin.bankroll(100, 900, 0.5)
  bankroll[i] <- hand
>>>>>>> b29b88f4562a784afd7a169bc3e74aab459d570e
}

mean(bankroll)
var(bankroll)
<<<<<<< HEAD
``` 
=======

```
>>>>>>> b29b88f4562a784afd7a169bc3e74aab459d570e
    e. the mean and variance of your bankroll after 500 hands (including busts).
```{r}
## set seed for replicability, running the simulation 
set.seed(1234)
## function for gambler's ruin, but now with 500 games
gamb.ruin.bankroll.500 <- function(bet, bankroll, p) {
  x <- 0 
  og.bet <- bet
  while(bankroll >= bet) { 
    game <- runif(1)
    if(game <= p) {
      bankroll = bankroll + bet
      bet = (bankroll+bet)*.10
    } else { 
      bankroll = bankroll - bet
      bet = og.bet
    }
  x <- x+1
    if(bankroll < bet){break} ##end if you don't have enough to bet
    if(x == 500){break} ##end if hit 500 games 
  }
  times.played <- x 
  return(bankroll)
}

bankroll <- rep(NA, length = n) ## set up empty bankroll 
simulation <- 5000
for(i in 1:simulation){
  hand <- gamb.ruin.bankroll.500(100, 900, 0.5)
  bankroll[i] <- hand
}

mean(bankroll)
var(bankroll)
```
Note: you *must* stop playing if your player has gone bust. How will you handle this in the `for` loop?

2. Repeat the previous problem with betting on black in American roulette, where the probability of winning on any spin is 18/38 for an even payout.
```{r}
set.seed(1234)
<<<<<<< HEAD
Gamble1 = function(bet,bankroll,p){
  x = 0
  og.bet <- bet
  while(bet > 0) {
    simulation = runif(1)
    if(simulation <= p){
      bankroll = bet + bankroll
      bet = (bet + bankroll)*0.1
    } else {
      bankroll = bankroll - bet
      bet = og.bet
      }
    x = x+1
    if(x ==500){
      break
      }
    if(bankroll < bet){
      break
      }
  }
  duration = x
  return(bankroll)
  
}

bankroll = numeric()
s = 5000
for(i in 1:s)
{
  next.bet = Gamble1(100,900,18/38)
  bankroll = c(bankroll, next.bet)
}
=======

gamb.ruin.bankroll.500 <- function(bet, bankroll, p) {
  x <- 0 
  og.bet <- bet
  while(bankroll >= bet) { 
    game <- runif(1)
    if(game <= p) {
      bankroll = bankroll + bet
      bet = (bankroll+bet)*.10
    } else { 
      bankroll = bankroll - bet
      bet = og.bet
    }
  x <- x+1
    if(bankroll < bet){break} 
    if(x == 500){break} 
  }
  times.played <- x 
  return(bankroll)
}

bankroll <- rep(NA, length = n)
simulation <- 5000
for(i in 1:simulation){
  hand <- gamb.ruin.bankroll.500(100, 900, 18/38) ## change probability 
  bankroll[i] <- hand
}

mean(bankroll)
var(bankroll)
```
>>>>>>> b29b88f4562a784afd7a169bc3e74aab459d570e

mean(bankroll)
var(bankroll)
```
3. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
    a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.
```{r}
<<<<<<< HEAD
Markov = function(bet, bankroll, p){
  x = 0
  prob <- p
  while(bet>0) {
    simulation = runif(1)
    if(simulation <= p){
      bankroll = bankroll +bet
      p = p+0.01
    } else {
      bankroll = bankroll-bet
      p = prob
    }
    x = x+1
    if(x ==100000){
      break
      }
    if(bankroll < bet){
      break
      }
  }
  return(bankroll)
}

size = numeric()
s = 100
for(i in 1:s)
{
  next.bet = Markov(100,900,0.48)
  size = c(size,next.bet)
}
mean(size)
print("Not a fair game")
```
    b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?
```{r}
set.seed(1)
prop = numeric()

for (x in 1:20){
  size = numeric()
  s = 100
for(i in 1:s){
  next.bet = Markov(100,900,0.46+0.002*x)
  size = c(size,next.bet)
}
  next.prop=length(which(size==0))/100
  prop = c(prop,next.prop)
}
prop
``` 
    c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 
```{r}
set.seed(1)
Markov = function(bet, bankroll, p, increment){
  x = 0
  prob <- p
  while(bet > 0) {
    simulation = runif(1)
    if(simulation <= p){
      bankroll = bankroll + bet
      p = p+increment
    } else {
      bankroll = bankroll - bet
      p = prob
    }
    x = x+1
    if(x == 100000){
      break
      }
    if(bankroll < bet){
      break
      }
  }
  return(bankroll)
}

prop = numeric()
for (x in 1:20){
  size = numeric()
  s = 100
for(i in 1:s){
  next.bet = Markov(100,900,0.48,0.002*x)
  size = c(size,next.bet)
}
  
  next.prop=length(which(size==0))/100
  prop=c(prop,next.prop)
}

prop
``` 
=======
## set seed for replicability, running the simulation 
set.seed(1234)
## basically gamb.ruin function with edits to p 
m.chains <- function(bet, bankroll, p) {
  x <- 0 
  chances <- p #create a vector for chance at winning at original p 
  while(bankroll >= bet) { 
    game <- runif(1)
    if(game <= p) { ## bets remain constant, p change 
      bankroll = bankroll + bet
      p = p + 0.01
    } else { ## if lose, p goes back to original 
      bankroll = bankroll - bet
      p = chances
    }
  x <- x+1
    if(bankroll < bet){break} ##end if you don't have enough to bet 
    if(x == 100000){break}
  }
  times.played <- x
  return(bankroll)
}

size.return <- rep(NA, length = n)
simulation <- 99
for(i in 1:simulation){
  hand <- m.chains(100, 900, 0.48)
  size.return[i] <- hand
}
mean(collect)
## game is not fair because mean of your returns are 0... 
```
    b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?
```{r}
set.seed(1234)
return <- rep(NA, length = n)
for (i in 1:15) { ## create a loop that runs m.chains but changes the beginning p everytime 
  return.size <- rep(NA, length = n) #empty vector for inputs from m.chains
  simulated <- 100
    for(x in 1:simulated) { 
      hand <- m.chains(100, 900, 0.48 + (0.002*i))
      return.size[x] <- hand
    }
  prop.return.0 <- length(
    which(return.size == 0)
  )/100 ## proportion of returns == 0, should be close to 50%
  return[i] <- prop.return.0
}
return

#when starting probability is 51% the game is most fair, slightly in favor of the gambler @ .49 
```
    c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 
```{r}
set.seed(1234)
## basically m.chains but with new arguement 
m.chains.increase <- function(bet, bankroll, p, p.increase) {
  x <- 0 
  chances <- p  
  while(bankroll >= bet) { 
    game <- runif(1)
    if(game <= p) {
      bankroll = bankroll + bet
      p = p + p.increase
    } else { ## if lose, p goes back to original 
      bankroll = bankroll - bet
      p = chances
    }
  x <- x+1
    if(bankroll < bet){break} ##end if you don't have enough to bet 
    if(x == 100000){break}
  }
  times.played <- x
  return(bankroll)
}
```

```{r}
## now to test things via loops
return <- rep(NA, length =n)
for (i in 1:15) { 
  return.size <- rep(NA, length =n)
  simulated <- 100
    for(x in 1:simulated) { ## change code so i now affects increment 
      hand <- m.chains.increase(100, 900, 0.48, .003*i)
      return.size[x] <- hand
    }
  prop.return.0 <- length(
    which(return.size == 0)
  )/100 ## proportion of returns == 0 
  return[i] <- prop.return.0
}
return
#somewhere between increments of .30 (56%) and .33 (46%) it is the most fair
```
>>>>>>> b29b88f4562a784afd7a169bc3e74aab459d570e

4. Creating a Bootstrap function. There is a particular concept called [bootstrapping]
(https://en.wikipedia.org/wiki/Bootstrapping_(statistics)) where we can easily create 95% confidence intervals, even for complex estimators.

The steps of this process are:

  a. Draw a sample, with replacement, from your data which is the same length of your data.
  b. Calculate the statistic of interest on this boostrap sample (ie mean, variance, regression,...)
  c. Peform steps 1:2 at least 1000 times over until you have a vector of your statistics. 
  d. The lower bound of a 95% CI will be the 0.025 percentile
  e. The upper bound of a 95% CI will be the 0.975 percentile

Make a function called `boot_ci` which calculates the 95% confidence interval in this manner. 

5. For problems 3b and 3c, you calculated a mean value. Because you saved these final results in a vector, use the bootstrap to estimate the variance of the return in each case for your final answer. Once you have these results, which game has the smaller variance in returns?